Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-05-21 ms.Rnw
blue box

\begin{abstract}
  We show how each member of a broad class of Markov population models uniquely induces a stochastic process on the space of genealogies.
  We then derive exact expressions for the likelihood of an observed genealogy in terms of filtering equations, which can be integrated using standard Monte Carlo methods.
  The class of population models that can be accommodated includes essentially all compartmental models.
\end{abstract}

@@

%% Population process induces Markov history and genealogy processes.
%% Using these, we derive equations for the likelihood of a genealogy conditional on the history.
%% We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
%% These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
%% We demonstrate with several examples.

@@

\appendix
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\renewcommand{\thefigure}{\thesection\arabic{figure}}
\renewcommand{\thetable}{\thesection\arabic{table}}

\section{Filter equations}
\label{sec:filter_eqns}

The likelihoods that appear in \cref{thm:pruned_uncond,thm:obsc_uncond} are integrals over large sets of histories.
As such, explicit expressions for them are not available, and we require mathematical tools to allow us to manipulate these quantities and devise algorithms for their numerical solution.
The \emph{filtering equations} we introduce here are suitable for these purposes, and we devote this appendix to exposing their essential properties.
This extremely convenient formalism has, to our knowledge, not been thoroughly exploited, though we note their resemblance to the constructions of \citet{Ogata1978}, \citet{Puri1986}, \citet{Kliemann1990}, and \citet{Giesecke2018}.

@@

\begin{lemma}
  \label{lemma:reg-filt}
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
  Let $\Vr_t$ be an $\Rp$-valued random process satisfying
  \begin{equation*}
    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{\mathclap{e\;\in\;\event{\H_t}}}{B(e,\Xt_e,\X_e)}.
  \end{equation*}
  Let the family of measures $\lambda_t$ on $\Xspace$ be defined by
  \begin{equation*}
    \lambda_t(\mathcal{E})=\Expect{\Vr_t\cdot\Indicator{\Xr_t\in{\mathcal{E}}}},
  \end{equation*}
  for measurable $\mathcal{E}$, and let $w(t,x)$ be the density of $\lambda_t$,
  \ie $\lambda_t(\dd{x})=w(t,x)\,\dd{x}$.
  In particular, $\Expect{\Vr_t}=\lambda_t(\Xspace)=\int{w(t,x)\,\dd{x}}$.
  Then $w$ satisfies the initial condition $w(0,x)=p_0(x)$ and the regular filter equation,
  \begin{equation}
    \label{eq:reg-filter-eq2}
    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
  \end{equation}
\end{lemma}
\begin{proof}
  Since $\Prob{\Vr_0=1}=1$, $\lambda_0(\mathcal{E})=\Prob{\Xr_0\in\mathcal{E}}$, which implies that $w(0,x)=p_0(x)$.
  For $t>0$ and $\Delta>0$ sufficiently small, the expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
  Accordingly, as $\Delta\downarrow{0}$,
  \begin{equation*}
    \begin{aligned}
      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
    \end{aligned}
  \end{equation*}
  In the limit, we obtain \cref{eq:reg-filter-eq2}, the regular filtering equation generated by $\alpha$, with boost $B$ and zero decay.
  %%  \AAK{[This depends on $\alpha$ and $B$ being continuous in their first arguments. We have assumed they are continuous almost everywhere.]}
\end{proof}

@@

\begin{table}[h!]
  \caption{
    \label{tab:sirs_model_elements}
    Elements of the SIRS model pertinent to the genealogy process.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cccl}
    \hline\hline
    $u$ & $\alpha_u$ & $r_u$ & Event type \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & 2 & pure birth \\
    $\lab{Recov}$  & $\gamma I$ \bigstrut & 0 & pure death \\
    $\lab{Wane}$   & $\omega R$ \bigstrut & 0 & neutral \\
    $\lab{Sample}$ & $\psi I$ \bigstrut & 1 & pure sample \\
    \hline\hline
  \end{tabular}
\end{table}

@@

\begin{table}[h!]
  \caption{
    \label{tab:seirs_model_elements}
    Elements of the SEIRS model pertinent to the genealogy process.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{ccccl}
    \hline\hline
    $u$ & $\alpha_u$ & $x\mapsto{x'}$ & $r_u$ & Event type \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & $(S,E)\mapsto(S-1,E+1)$  \bigstrut & $(1,1)$ & pure birth \\
    $\lab{Prog}$   & $\sigma E$ \bigstrut & $(E,I)\mapsto(E-1,I+1)$ \bigstrut & $(0,1)$ & pure migration \\
    $\lab{Recov}$  & $\gamma I$ \bigstrut & $(I,R)\mapsto(I-1,R+1)$ \bigstrut & $(0,0)$ & pure death \\
    $\lab{Wane}$   & $\omega R$ \bigstrut & $(S,R)\mapsto(S+1,R-1)$ \bigstrut & $(0,0)$ & neutral \\
    $\lab{Sample}$ & $\psi I$ \bigstrut & $x\mapsto{x}$ \bigstrut & $(0,1)$ & pure sample \\
    \hline\hline
  \end{tabular}
\end{table}

\begin{table}[h!]
  \caption{
    \label{tab:seirs_integ}
    Elements of a scheme for computing the likelihood under the SEIRS model.
    The $y\mapsto{y'}$ column depicts the proposed painting schematically.
    Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{c c c c c c c c | c}
    \hline\hline
    & $u$ & $s$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & n & $Q$ & Line\\[2ex]
    \hline
    \multirow[c]{10}{*}[-1ex]{\rotatebox{90}{$t\notin{{B}\cup{S_0}\cup{S_1}}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & $(0,0)$ & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I+1-\ell_I}{I+1}$ & 1 & 1 & 1 \\
    \cline{3-8}
    && $(1,0)$ & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 2 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 3 \\
    \cline{3-8}
    && $(1,1)$ & --- & --- & 0 & 0 & 0 & 4 \\
    \cline{2-8}
    &\multirow{2}{*}{$\lab{Prog}$}
    & $(0,0)$ & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E+1-\ell_E}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 1 & 5 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 1 & 6 \\
    \cline{2-8}
    & $\lab{Recov}$ & $(0,0)$ & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & 1 & 1 & 7 \\
    \cline{2-8}
    & $\lab{Wane}$ & $(0,0)$ & 1 \bigstrut & \unbroken \bigstrut & 1 & 1 & 1 & 8 \\
    \cline{2-8}
    & $\lab{Sample}$ & $\ast$ & --- & --- & 1 & 0 & 0 & 9 \\
    \cline{1-8}
    \multirow[c]{4}{*}[-3ex]{\rotatebox{90}{$t\in{B}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & \multirow{4}{*}{$(1,1)$}
    & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 10 \\
    \cline{4-8}
    &&& $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 11 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchupnix \bigstrut & 0 & 0 & 0 & 12 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchdownnix \bigstrut & 0 & 0 & 0 & 13 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- \bigstrut & --- & 0 & 0 & 0 & 14 \\
    \cline{1-8}
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_0}$} \bigstrut}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,0)$}
    & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & 1 & 1 & 1 & 15 \\
    \cline{4-8}
    &&& --- & \samplezeronix \bigstrut & 0 & 0 & 0 & 16 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 17 \\
    \cline{1-8}
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_1}$}}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,1)$}
    & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & 1 & 1 & 1 & 18 \\
    \cline{4-8}
    &&& --- & \sampleonenix \bigstrut & 0 & 0 & 0 & 19 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 20 \\
    \hline\hline
  \end{tabular}
\end{table}

A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
Given a particular choice of importance-sampling kernel, the filtering equation follows.
For example, with the choices as in \cref{tab:seirs_integ}, the filtering equation is as follows.
As previously, let $B$ be the set of branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, we have
\begin{mathsize}{8pt}{9pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{w}}{\partial{t}}(t,S,E,I,R,y)=
      %% &\frac{\beta\,(S+1)\,I}{N}\,\frac{I-\ell_I}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,y)\\
      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{I-\ell_I}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_2)\\
      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_3)
      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
      %% &+\sigma\,(E+1)\frac{E+1-\ell_E}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{I-\ell_I}{I}\,\frac{E+1}{E+1-\ell_E}\,w(t,S,E+1,I-1,R,y)\\
      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\frac{1}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{E+1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6)
      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
      %% -\omega\,R\,w(t,S,E,I,R,y)
      %% -\psi\,I\,w(t,S,E,I,R,y).
      %%
      &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_I}{I}\right)\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
      &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)\\
      &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_3^k)
      -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
      &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
      &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
      -\sigma\,E\,w(t,S,E,I,R,y)\\
      &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
      -\gamma\,I\,w(t,S,E,I,R,y)\\
      &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
      -\omega\,R\,w(t,S,E,I,R,y)
      -\psi\,I\,w(t,S,E,I,R,y).
      %%
      %% &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)
      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
      %% &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
      %% -\omega\,R\,w(t,S,E,I,R,y)
      %% -\psi\,I\,w(t,S,E,I,R,y).
    \end{split}
  \end{equation*}
\end{mathsize}%

>>>>

\begin{abstract}
  We show how each member of a broad class of Markov population models uniquely induces a stochastic process on the space of genealogies.
  We then derive exact expressions for the likelihood of an observed genealogy in terms of filter equations, which can be integrated using standard Monte Carlo methods.
  The class of population models that can be accommodated includes essentially all compartmental models.
\end{abstract}

@@

%% Population process induces Markov history and genealogy processes.
%% Using these, we derive equations for the likelihood of a genealogy conditional on the history.
%% We then integrate out the history to obtain nonlinear filter equations, the solution of which yields the likelihood.
%% These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
%% We demonstrate with several examples.

@@

\appendix
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
%%\setcounter{tcbcounter}{0}

\titleformat{\section}[hang]{\large\bfseries}{Appendix \periodafter\thesection}{2ex}{\periodafter}{}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\renewcommand{\thefigure}{\thesection\arabic{figure}}
\renewcommand{\thetable}{\thesection\arabic{table}}
%%\renewcommand{\thetcbcounter}{\thesection\arabic{tcbcounter}}

\section{filter equations}
\label{sec:filter_eqns}

The likelihoods that appear in \cref{thm:pruned_uncond,thm:obsc_uncond} are integrals over large sets of histories.
As such, explicit expressions for them are not available, and we require mathematical tools to allow us to manipulate these quantities and devise algorithms for their numerical solution.
The \emph{filter equations} we introduce here are suitable for these purposes, and we devote this appendix to exposing their essential properties.
This extremely convenient formalism has, to our knowledge, not been thoroughly exploited, though we note their resemblance to the constructions of \citet{Ogata1978}, \citet{Puri1986}, \citet{Kliemann1990}, and \citet{Giesecke2018}.

@@

\begin{lemma}
  \label{lemma:reg-filt}
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
  Let $\Vr_t$ be an $\Rp$-valued random process satisfying
  \begin{equation*}
    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{\mathclap{e\;\in\;\event{\H_t}}}{B(e,\Xt_e,\X_e)}.
  \end{equation*}
  Let the family of measures $\lambda_t$ on $\Xspace$ be defined by
  \begin{equation*}
    \lambda_t(\mathcal{E})=\Expect{\Vr_t\cdot\Indicator{\Xr_t\in{\mathcal{E}}}},
  \end{equation*}
  for measurable $\mathcal{E}$, and let $w(t,x)$ be the density of $\lambda_t$,
  \ie $\lambda_t(\dd{x})=w(t,x)\,\dd{x}$.
  In particular, $\Expect{\Vr_t}=\lambda_t(\Xspace)=\int{w(t,x)\,\dd{x}}$.
  Then $w$ satisfies the initial condition $w(0,x)=p_0(x)$ and the regular filter equation,
  \begin{equation}
    \label{eq:reg-filter-eq2}
    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
  \end{equation}
\end{lemma}
\begin{proof}
  Since $\Prob{\Vr_0=1}=1$, $\lambda_0(\mathcal{E})=\Prob{\Xr_0\in\mathcal{E}}$, which implies that $w(0,x)=p_0(x)$.
  For $t>0$ and $\Delta>0$ sufficiently small, the expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
  Accordingly, as $\Delta\downarrow{0}$,
  \begin{equation*}
    \begin{aligned}
      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
    \end{aligned}
  \end{equation*}
  In the limit, we obtain \cref{eq:reg-filter-eq2}, the regular filter equation generated by $\alpha$, with boost $B$ and zero decay.
  %%  \AAK{[This depends on $\alpha$ and $B$ being continuous in their first arguments. We have assumed they are continuous almost everywhere.]}
\end{proof}

@@

\begin{table}[h!]
  \caption{
    \label{tab:sirs_model_elements}
    Elements of the SIRS model pertinent to the genealogy process.
    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r_u$), and event type for each of the model's four marks ($u$).
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{ccccl}
    \hline\hline
    $u$ & $\alpha_u$ \bigstrut & $x\mapsto{x'}$ \bigstrut & $r_u$ \bigstrut & Event type \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & $(S,I)\mapsto(S-1,I+1)$ \bigstrut & 2 & pure birth \\
    $\lab{Recov}$  & $\gamma I$ \bigstrut & $(I,R)\mapsto(I-1,R+1)$ \bigstrut & 0 & pure death \\
    $\lab{Wane}$   & $\omega R$ \bigstrut & $(S,R)\mapsto(S+1,R-1)$ \bigstrut & 0 & neutral \\
    $\lab{Sample}$ & $\psi I$ \bigstrut & $x\mapsto{x}$ \bigstrut & 1 & pure sample \\
    \hline\hline
  \end{tabular}
\end{table}

@@

\begin{table}[h!]
  \caption{
    \label{tab:seirs_model_elements}
    Elements of the SEIRS model pertinent to the genealogy process.
    The table shows the rate ($\alpha_u$), jump ($x\mapsto{x'}$), production ($r_u$), and event type for each of the model's five marks ($u$).
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{ccccl}
    \hline\hline
    $u$ & $\alpha_u$ & $x\mapsto{x'}$ & $r_u$ & Event type \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & $(S,E)\mapsto(S-1,E+1)$  \bigstrut & $(1,1)$ & pure birth \\
    $\lab{Prog}$   & $\sigma E$ \bigstrut & $(E,I)\mapsto(E-1,I+1)$ \bigstrut & $(0,1)$ & pure migration \\
    $\lab{Recov}$  & $\gamma I$ \bigstrut & $(I,R)\mapsto(I-1,R+1)$ \bigstrut & $(0,0)$ & pure death \\
    $\lab{Wane}$   & $\omega R$ \bigstrut & $(S,R)\mapsto(S+1,R-1)$ \bigstrut & $(0,0)$ & neutral \\
    $\lab{Sample}$ & $\psi I$ \bigstrut & $x\mapsto{x}$ \bigstrut & $(0,1)$ & pure sample \\
    \hline\hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{
    \label{tab:seirs_integ}
    Elements of a scheme for numerically computing the likelihood under the SEIRS model.
    The regular portion of the filter equation holds in between genealogical events (\ie for $t\notin{{B}\cup{S_0}\cup{S_1}}$);
    the singular portion describes the effect of these events ($t\in{{B}\cup{S_0}\cup{S_1}}$).
    Each line corresponds to a potential event, but only those for which $Q=1$ appear in the equation.
    An event with mark $u$ and saturation $s$ has the boost given by the binomial ratio shown (third column).
    The $y\mapsto{y'}$ column depicts the proposed painting schematically, and $\pi$ is the probability of that proposal.
    Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
    The filter equation contains $n$ terms of each type.
    An asterisk ($\ast$) stands for cases not explicitly mentioned.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{c c c c c c c c | c}
    \hline\hline
    & $u$ & $s$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & n & $Q$ & Line\\[2ex]
    \hline
    \multirow[c]{10}{*}[-1ex]{\rotatebox{90}{$t\notin{{B}\cup{S_0}\cup{S_1}}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & $(0,0)$ & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I+1-\ell_I}{I+1}$ & 1 & 1 & 1 \\
    \cline{3-8}
    && $(1,0)$ & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 2 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 3 \\
    \cline{3-8}
    && $(1,1)$ & --- & --- & 0 & 0 & 0 & 4 \\
    \cline{2-8}
    &\multirow{2}{*}{$\lab{Prog}$}
    & $(0,0)$ & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E+1-\ell_E}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 1 & 5 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 1 & 6 \\
    \cline{2-8}
    & $\lab{Recov}$ & $(0,0)$ & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & 1 & 1 & 7 \\
    \cline{2-8}
    & $\lab{Wane}$ & $(0,0)$ & 1 \bigstrut & \unbroken \bigstrut & 1 & 1 & 1 & 8 \\
    \cline{2-8}
    & $\lab{Sample}$ & $\ast$ & --- & --- & 1 & 0 & 0 & 9 \\
    \cline{1-8}
    \multirow[c]{4}{*}[-3ex]{\rotatebox{90}{$t\in{B}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & \multirow{4}{*}{$(1,1)$}
    & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 10 \\
    \cline{4-8}
    &&& $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 11 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchupnix \bigstrut & 0 & 0 & 0 & 12 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchdownnix \bigstrut & 0 & 0 & 0 & 13 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- \bigstrut & --- & 0 & 0 & 0 & 14 \\
    \cline{1-8}
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_0}$} \bigstrut}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,0)$}
    & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & 1 & 1 & 1 & 15 \\
    \cline{4-8}
    &&& --- & \samplezeronix \bigstrut & 0 & 0 & 0 & 16 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 17 \\
    \cline{1-8}
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_1}$}}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,1)$}
    & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & 1 & 1 & 1 & 18 \\
    \cline{4-8}
    &&& --- & \sampleonenix \bigstrut & 0 & 0 & 0 & 19 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 20 \\
    \hline\hline
  \end{tabular}
\end{table}

The filter equation corresponding to the scheme of \cref{tab:seirs_integ} is presented in \Cref{box:SEIRS}.
Some numerical results are presented in \cref{fig:seirs_example}.

\clearpage
\newpage

\begin{bluebox}[label={box:SEIRS}]{Filter equation for the SEIRS model}
  As previously, given an obscured genealogy, let $B$ be the set of its branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
  Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, the filter equation reads:
  \begin{mathsize}{9pt}{10pt}
    \begin{equation*}
      \begin{split}
        \frac{\partial{w}}{\partial{t}}(t,S,E,I,R,y)=&
        \frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_E}{E}\right)\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,y)\\
        &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)\\
        &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_3^k)
        -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
        &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
        &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
        -\sigma\,E\,w(t,S,E,I,R,y)\\
        &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
        -\gamma\,I\,w(t,S,E,I,R,y)\\
        &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
        -\omega\,R\,w(t,S,E,I,R,y)
        -\psi\,I\,w(t,S,E,I,R,y).
      \end{split}
    \end{equation*}
  \end{mathsize}%
  Here, $\leftlim{y}_j^k$ refers to the coloring of the tree immediately preceding the proposal indicated on line $j$ of \cref{tab:seirs_integ}.
  The integer $k$ specifies the particular branch on which the change occurs.

  The singular portion of the filter equation has one component for each distinct type of genealogical event:
  \begin{mathsize}{9pt}{10pt}
    \begin{equation*}
      w(t,S,E,I,R,y)=
      \begin{cases}
        \frac{\beta\,(S+1)\,I}{N}\,\frac{1}{E I}\,\wt(t,S+1,E-1,I,R,\leftlim{y}_{10})\\[2ex]
        \qquad+\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{E I}\,\wt(t,S+1,E-1,I,R,\leftlim{y}_{11}), & t\in{B},\\[2ex]
        \psi\,\left(I-\ell_I\right)\,\wt(t,S,E,I,R,\leftlim{y}_{15}), & t\in{S_0},\\[2ex]
        \psi\,\wt(t,S,E,I,R,\leftlim{y}_{18}), & t\in{S_1}.
      \end{cases}
    \end{equation*}
  \end{mathsize}%

  A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
  With a particular choice of importance-sampling kernel, the filter equation uniquely specifies a sequential Monte Carlo algorithm for estimating the likelihood.
  The choices made in \cref{tab:seirs_integ} underlie the results displayed in \cref{fig:seirs_example}.
\end{bluebox}
